### 2024-02-16 doc/15yrsChange.Rmd
General Notes and Pests/Pathogens Outline

@@

White-tailed deer (Odocoileus virginianus) represent an endemic nuisance species prevalent in eastern deciduous forests. Their populations have substantially increased in the past 50 years (McShea et al. 1997), due in large part to human influence. (Found in McGarvey et al. 2013 – Brown et al. 2000, Côté et al. 2004, Rooney 2001).  White-tailed deer preferentially browse on woody herbaceous species in their earliest life stages (McGarvey et al. 2013), with overabundant herbivory negatively impacting seedling and sapling survival, growth and density (Found in McGarvey et al. 2013 – Dzieciolowski 1980, Gill and Beardall 2001, Healy 1997, Konig 1976, Putman et al. 1989).  Models predict that chronic overabundance will reduce understory diversity and decrease the abundance of traditionally dominant species (Holm et al. 2013).  Due to climate change and human activity, many non-endemic plant species are being introduced or number in forest ecosystems (Reference).  Deer find many of these species, including pawpaw (Asimina triloba), to be unpalatable (Found in McGarvey 2013 – Asnani 2006), and do not consume them at the same rates relative to their native counterparts (Reference), enabling them to form dense stands in forest understories (Found in Knauer et al. 2023 – Horsley and Marquis 1983; Stromayer and Warren 1997; Royo and Carson 2006).  This lack of browsing pressure, coupled with characteristics such as fast growth rate and greater adaptability to altered conditions precipitated by climate change, allow nuisance plant species to outcompete other species in the understory (Reference). The interaction of these nuisance species on the landscape contributes to a recruitment failure in forests, [*Where individuals from certain tree species do not enter the mature population - Do we need to define recruitment failure, or is that considered common knowledge?  If so, reference*).  The extent of the consequences derived from recruitment failure on species composition and forest structure is often not apparent for years (McGarvey et al. 2013)

@@

*What is the best way to describe the forest type in our study site? I've seen "mixed mature deciduous forest", "mature secondary eastern deciduous forest", and "mesic temperate deciduous forest".  I want to make sure I'm being consistent.*

@@

The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38 53’36.6“N, 78 08’43.4 “W).  The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of secondary eastern mixed deciduous forest. Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*) and witch hazel (*Hamamelis virginiana*) (SCBI site description). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al. 2013).  Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).  The plot, which includes a 4-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the World.

@@

A number of surveys are regularly conducted at the site, two of which are included in our analysis.  As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).  This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter at 1.3 meters in height (referred to as diameter at breast height or dbh).  The census records information regarding the dbh, species, status and location of each stem included in the survey.  Each plant is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field. For multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.  The location of each individual within its respective 20 x 20 meter quadrat is recorded on a map of the plot on ArcGIS FieldMaps.  Established in 2008, there have been 4 censuses at the site, comprising 20 years of detailed forest dynamics data. In addition to the ForestGEO census, an annual mortality survey is conducted on the site.   Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.  Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.

*Is dbh considered common knowledge? Does it need to be defined?*

*Check with Luca to use consistent terminology here and in the analysis portion of the methods for ID tag, stem tag, etc.*

>>>>

Two Paragraphs

1. The global implications of pests and pathogens
2. History of pests and pathogens in eastern deciduous forests
  1. Chestnut blight
  2. Dutch elm disease
  3. Hemlock Wooly Adelgid
  4. Emerald Ash-borer

@@

White-tailed deer (*Odocoileus virginianus*) are a nuisance species prevalent in eastern deciduous forests.*Start with a stronger sentence about the impact of deer on herbivory - Ex. White-tailed deer are the dominant herbivore in eastern deciduous forests (Reference)* Their populations have dramatically increased in the past 50 years (McShea et al. 1997).  *Find a reference with a specific number of increase in deer populations* due in large part to global change drivers. (Found in McGarvey et al. 2013 – Brown et al. 2000, Côté et al. 2004, Rooney 2001). White-tailed deer preferentially browse on woody plant species in their earliest life stages (McGarvey et al. 2013), with overabundant herbivory reduce seedling and sapling survival, growth and density (Found in McGarvey et al. 2013 – Dzieciolowski 1980, Gill and Beardall 2001, Healy 1997, Konig 1976, Putman et al. 1989).   Chronic overabundance has been shown to reduce understory diversity and decrease the abundance of traditionally dominant species (*NEW CITATION*).  Due to climate change and human activity, many non-endemic plant species are being introduced or number in forest ecosystems (Reference).  Deer find many of these species, including pawpaw (Asimina triloba), to be unpalatable (Found in McGarvey 2013 – Asnani 2006), and do not consume them at the same rates relative to their native counterparts (Reference), enabling them to form dense stands in forest understories (Found in Knauer et al. 2023 – Horsley and Marquis 1983; Stromayer and Warren 1997; Royo and Carson 2006).  This lack of browsing pressure, coupled with characteristics such as fast growth rate and greater adaptability to altered conditions precipitated by climate change, allow nuisance plant species to outcompete other species in the understory (Reference). The interaction of these nuisance species on the landscape contributes to a recruitment failure in forests, [*Where individuals from certain tree species do not enter the mature population - Do we need to define recruitment failure, or is that considered common knowledge?  If so, reference*).  The extent of the consequences derived from recruitment failure on species composition and forest structure is often not apparent for years (McGarvey et al. 2013)

@@

@@

The study was conducted in the 25.6-hectare Large Forest Dynamics plot at the National Zoo and Conservation Biology Institute in Front Royal, Virginia (38 53’36.6“N, 78 08’43.4 “W).  The plot, which is located in the central Appalachian Mountains adjacent to Shenandoah National Park, is composed of mature secondary eastern deciduous forest. Situated in the Appalachian Oak forest region, the canopy is dominated by tulip poplar (*Liriodendron tulipifera*), oak (*Quercus spp.*), and hickory (*Carya spp.*) and the understory is primarily composed of spicebush (*Lindera benzoin*), paw-paw (*Asimina triloba*), American hornbeam (*Carpinus caroliniana*) and witch hazel (*Hamamelis virginiana*) (SCBI site description). The land-use history of the site is varied, including periods of agricultural development and intensive logging, with dendrological data estimating canopy tree establishment around 1900 (SI archives, Bourg et al. 2013).  Natural disturbances at the plot consist of wind and infrequent ice storms (Anderson-Teixiera 2015).  The plot, which includes a 4-hectare deer exclosure that has decreased the presence of deer since 1990, is divided into 640 quadrats, each measuring 20 x 20 meters. It is one of 78 sites in the Forest Global Earth Observatory (ForestGEO), a global network of forest dynamic plots that promotes comparative forest ecology studies around the World.

*Add information about the history of pests and pathogens at this site and at similar forests*

@@

A number of surveys are regularly conducted at the site, two of which are included in our analysis.  As part of the ForestGEO network, the plot undergoes a comprehensive woody plant inventory every five years, according to the protocol outlined in Condit (1998).  This inventory, hereby referred to as the census, includes all stems greater than 1 centimeter in diameter breast height (1.3 meters; hereafter referred to as dbh).  The census records information regarding the dbh, species, status and location of each stem included in the survey.  Each plant is assigned an identifying number for sequential data collection in subsequent censuses and outfitted with a metal tag in the field. For multi-stemmed individuals, each additional stem that surpasses the 1 cm dbh threshold receives a stem number and associated tag.  The location of each individual within its respective 20 x 20 meter quadrat is recorded on a map of the plot on ArcGIS FieldMaps.  Established in 2008, there have been 4 censuses at the site, comprising 20 years of detailed forest dynamics data. In addition to the ForestGEO census, an annual mortality survey is conducted on the site.   Through assessing individual tree health, this study illustrates trends in forest mortality and identifies factors that are associated with death.  Data is collected on the current status, canopy position, percentage of crown living and intact, and visible indicators of poor tree health, such as physical damage, potential pathogens and insect infestation.

*Add Condit 1998 in references*

*In response to the presence of Emerald Ashborer, a mortality survey was initiated, then improved in year [X].
